
\mysectionnn{Bibliography}

{
 Albert, David Z. \textit{Quantum Mechanics and Experience}.
Harvard University Press, 1994.}

{
 Alexander, Scott. ``Why I Am Not Rene
Descartes.'' \textit{Slate Star Codex (blog)} (2014).
\url{http://slatestarcodex.com/2014/11/27/why-i-am-not-rene-descartes/}.}

{
 Allais, Maurice. ``Le Comportement de
l'Homme Rationnel devant le Risque: Critique des
Postulats et Axiomes de l'Ecole
Americaine.'' \textit{Econometrica} 21, no. 4 (1953):
503--546. doi:10.2307/1907921.}

{
 Ariely, Dan. \textit{Predictably Irrational: The Hidden Forces
That Shape Our Decisions}. HarperCollins, 2008.}

{
 Asch, Solomon E. ``Studies of Independence and
Conformity: A Minority of One Against a Unanimous
Majority.'' \textit{Psychological Monographs} 70
(1956).}

{
 Ashmore, Richard D., Vasantha Ramchandra, and Russell A. Jones.
``Censorship as an Attitude Change
Induction.'' \textit{Paper presented at Eastern
Psychological Association meeting} (1971).}

{
 Asimov, Isaac. \textit{The Relativity of Wrong}. Oxford University
Press, 1989.}

{
 Baez, John. ``The Crackpot
Index.'' 1998.
\url{http://math.ucr.edu/home/baez/crackpot.html}.}

{
 Banks, Iain. \textit{The Player of Games}. Orbit, 1989.}

{
 Baratz, Daphna. \textit{How Justified Is the
``Obvious'' Reaction?} Stanford
University, 1983.}

{
 Barbour, Julian. \textit{The End of Time: The Next Revolution in
Physics}. 1st ed. New York: Oxford University Press, 1999.}

{
 Baron, Jonathan. \textit{Thinking and Deciding}. Cambridge
University Press, 2007.}

{
 Baron, Jonathan, and Joshua D. Greene.
``Determinants of Insensitivity to Quantity in
Valuation of Public Goods: Contribution, Warm Glow, Budget Constraints,
Availability, and Prominence.'' \textit{Journal of
Experimental Psychology: Applied} 2, no. 2 (1996): 107--125.
doi:10.1037/1076-898X.2.2.107.}

{
 Barrett, Jeffrey. \textit{Everett's Relative-State
Formulation of Quantum Mechanics}. Edited by Edward N. Zalta.
\url{http://plato.stanford.edu/archives/fall2008/entries/qm-everett/}.}

{
 Benson, Peter L., Stuart A. Karabenick, and Richard M. Lerner.
``Pretty Pleases: The Effects of Physical
Attractiveness, Race, and Sex on Receiving Help.''
\textit{Journal of Experimental Social Psychology} 12 (5 1976):
409--415. doi:10.1016/0022-1031(76)90073-1.}

{
 Bond, Rod, and Peter B. Smith. ``Culture and
Conformity: A Meta-Analysis of Studies Using Asch's
(1952b, 1956) Line Judgment Task.''
\textit{Psychological Bulletin} 119 (1996): 111--137.}

{
 Bostrom, Nick. ``A History of Transhumanist
Thought.'' \textit{Journal of Evolution and
Technology} 14, no. 1 (2005): 1--25.
\url{http://www.nickbostrom.com/papers/history.pdf}.}

{
 {}---{}---{}---. ``Existential Risks: Analyzing
Human Extinction Scenarios and Related Hazards.''
\textit{Journal of Evolution and Technology} 9 (2002).
\url{http://www.jetpress.org/volume9/risks.html}.}

{
 {}---{}---{}---. \textit{Superintelligence: Paths, Dangers,
Strategies}. Oxford University Press, 2014.}

{
 Bostrom, Nick, and Milan M. \'Cirkovi\'c, eds. \textit{Global
Catastrophic Risks}. New York: Oxford University Press, 2008.}

{
 Bostrom, Nick, and Julian Savulescu. ``Human
Enhancement Ethics: The State of the Debate.'' In
\textit{Human Enhancement}, edited by Nick Bostrom and Julian
Savulescu. 2009.}

{
 Bostrom, Nick, and Eliezer Yudkowsky. ``The
Ethics of Artificial Intelligence.'' In \textit{The
Cambridge Handbook of Artificial Intelligence}, edited by Keith
Frankish and William Ramsey. New York: Cambridge University Press,
2014.}

{
 Bourget, David, and David J. Chalmers. ``What Do
Philosophers Believe?'' \textit{Philosophical
Studies} (2013): 1--36.}

{
 Boynton, Robert S. ``The Birth of an Idea: A
Profile of Frank Sulloway.'' \textit{The New Yorker}
(October 1999).}

{
 Brehm, Sharon S., and Marsha Weintraub.
``Physical Barriers and Psychological Reactance:
Two-year-olds' Responses to Threats to
Freedom.'' \textit{Journal of Personality and Social
Psychology} 35 (1977): 830--836.}

{
 Broeder, Dale. ``The University of Chicago Jury
Project.'' \textit{Nebraska Law Review} 38 (1959):
760--774.}

{
 Brown, Kevin. \textit{Reflections On Relativity}. Raleigh, NC:
printed by author, 2011. \url{http://www.mathpages.com/rr/rrtoc.htm}.}

{
 Brust, Steven. \textit{The Paths of the Dead}. Vol. 1 of The
Viscount of Adrilankha. Tor Books, 2002.}

{
 Budesheim, Thomas Lee, and Stephen DePaola.
``Beauty or the Beast?: The Effects of Appearance,
Personality, and Issue Information on Evaluations of Political
Candidates.'' \textit{Personality and Social
Psychology Bulletin} 20 (4 1994): 339--348.
doi:10.1177/0146167294204001.}

{
 Buehler, Roger, Dale Griffin, and Michael Ross.
``Exploring the `Planning
Fallacy': Why People Underestimate Their Task
Completion Times.'' \textit{Journal of Personality
and Social Psychology} 67, no. 3 (1994): 366--381.
doi:10.1037/0022-3514.67.3.366.}

{
 {}---{}---{}---. ``Inside the Planning Fallacy:
The Causes and Consequences of Optimistic Time
Predictions.'' In Gilovich, Griffin, and Kahneman,
\textit{Heuristics and Biases}, 250--270.}

{
 {}---{}---{}---. ``It's About
Time: Optimistic Predictions in Work and Love.''
\textit{European Review of Social Psychology} 6, no. 1 (1995): 1--32.
doi:10.1080/14792779343000112.}

{
 Bujold, Lois McMaster. \textit{Komarr}. Miles Vorkosigan
Adventures. Baen, 1999.}

{
 Burton, Ian, Robert W. Kates, and Gilbert F. White. \textit{The
Environment as Hazard}. 1st ed. New York: Oxford University Press,
1978.}

{
 Campbell, Richmond, and Lanning Snowden, eds. \textit{Paradoxes of
Rationality and Cooperation: Prisoner's Dilemma and
Newcomb's Problem}. Vancouver: University of British
Columbia Press, 1985.}

{
 Carson, Richard T., and Robert Cameron Mitchell.
``Sequencing and Nesting in Contingent Valuation
Surveys.'' \textit{Journal of Environmental Economics
and Management} 28, no. 2 (1995): 155--173.
doi:10.1006/jeem.1995.1011.}

{
 Casscells, Ward, Arno Schoenberger, and Thomas Graboys.
``Interpretation by Physicians of Clinical Laboratory
Results.'' \textit{New England Journal of Medicine}
299 (1978): 999--1001.}

{
 Castellow, Wilbur A., Karl L. Wuensch, and Charles H. Moore.
``Effects of Physical Attractiveness of the Plaintiff
and Defendant in Sexual Harassment Judgments.''
\textit{Journal of Social Behavior and Personality} 5 (6 1990):
547--562.}

{
 Chaiken, Shelly. ``Communicator Physical
Attractiveness and Persuasion.'' \textit{Journal of
Personality and Social Psychology} 37 (8 1979): 1387--1397.
doi:10.1037/0022-3514.37.8.1387.}

{
 Chalmers, David J. \textit{The Conscious Mind: In Search of a
Fundamental Theory}. New York: Oxford University Press, 1996.}

{
 Chapman, Graham, et al. \textit{Monty Python's The
Life of Brian (of Nazareth)}. Eyre Methuen, 1979.}

{
 Chapman, Gretchen B., and Eric J. Johnson.
``Incorporating the Irrelevant: Anchors in Judgments
of Belief and Value.'' In Gilovich, Griffin, and
Kahneman, \textit{Heuristics and Biases}, 120--138.}

{
 Cherryh, Carolyn J. \textit{The Paladin}. Baen, 2002.}

{
 Cialdini, Robert B. \textit{Influence: Science and Practice}.
Boston: Allyn \& Bacon, 2001.}

{
 {}---{}---{}---. \textit{Influence: The Psychology of Persuasion:
Revised Edition}. New York: Quill, 1993.}

{
 Cleaver, Jerry. \textit{Immediate Fiction: A Complete Writing
Course}. Macmillan, 2004.}

{
 Combs, Barbara, and Paul Slovic. ``Newspaper
Coverage of Causes of Death.'' \textit{Journalism \&
Mass Communication Quarterly} 56, no. 4 (1979): 837--849.
doi:10.1177/107769907905600420.}

{
 Crawford, Charles B., Brenda E. Salter, and Kerry L. Jang.
``Human Grief: Is Its Intensity Related to the
Reproductive Value of the Deceased?''
\textit{Ethology and Sociobiology} 10, no. 4 (1989): 297--307.}

{
 Darwin, Charles. \textit{On the Origin of Species by Means of
Natural Selection; or, The Preservation of Favoured Races in the
Struggle for Life}. 1st ed. London: John Murray, 1859.
\url{http://darwin-online.org.uk/content/frameset?viewtype=text\&itemID=F373\&pageseq=1}.}

{
 {}---{}---{}---. \textit{The Descent of Man, and Selection in
Relation to Sex}. 2nd ed. London: John Murray, 1874.
\url{http://darwin-online.org.uk/content/frameset?itemID=F944\&viewtype=text\&pageseq=1}.}

{
 Darwin, Francis, ed. \textit{The Life and Letters of Charles
Darwin}. John Murray, 1887.}

{
 Dawes, Robyn M. \textit{House of Cards: Psychology and
Psychotherapy Built on Myth}. Free Press, 1996.}

{
 {}---{}---{}---. \textit{Rational Choice in An Uncertain World}.
1st ed. Edited by Jerome Kagan. San Diego, CA: Harcourt Brace
Jovanovich, 1988.}

{
 De Camp, Lyon Sprague, and Fletcher Pratt. \textit{The Incomplete
Enchanter}. New York: Henry Holt \& Company, 1941.}

{
 Denes-Raj, Veronika, and Seymour Epstein.
``Conflict between Intuitive and Rational Processing:
When People Behave against Their Better Judgment.''
\textit{Journal of Personality and Social Psychology} 66 (5 1994):
819--829. doi:10.1037/0022-3514.66.5.819.}

{
 Dennett, Daniel C. \textit{Breaking the Spell: Religion as a
Natural Phenomenon}. Penguin, 2006.}

{
 {}---{}---{}---. \textit{Darwin's Dangerous Idea:
Evolution and the Meanings of Life}. Simon \& Schuster, 1995.}

{
 {}---{}---{}---. \textit{Freedom Evolves}. Viking Books, 2003.}

{
 {}---{}---{}---. ``The Unimagined
Preposterousness of Zombies.'' \textit{Journal of
Consciousness Studies} 2 (4 1995): 322--26.}

{
 Descartes, René. \textit{Discours de la Méthode}. Librairie des
Bibliophiles, 1887.}

{
 Desvousges, William H., F. Reed Johnson, Richard W. Dunford, Kevin
J. Boyle, Sara P. Hudson, and K. Nicole Wilson. \textit{Measuring
Nonuse Damages Using Contingent Valuation: An Experimental Evaluation
of Accuracy}. Technical report. Research Triangle Park, NC: RTI
International, 2010. doi:10.3768/rtipress.2009.bk.0001.1009.}

{
 Downs, A. Chris, and Phillip M. Lyons. ``Natural
Observations of the Links Between Attractiveness and Initial Legal
Judgments.'' \textit{Personality and Social
Psychology Bulletin} 17 (5 1991): 541--547.
doi:10.1177/0146167291175009.}

{
 Drescher, Gary L. \textit{Good and Real: Demystifying Paradoxes
from Physics to Ethics}. Cambridge, MA: MIT Press, 2006.}

{
 Eagly, Alice H., Richard D. Ashmore, Mona G. Makhijani, and Laura
C. Longo. ``What Is Beautiful Is Good, But \ldots A
Meta-analytic Review of Research on the Physical Attractiveness
Stereotype.'' \textit{Psychological Bulletin} 110 (1
1991): 109--128. doi:10.1037/0033-2909.110.1.109.}

{
 Eddy, David M. ``Probabilistic Reasoning in
Clinical Medicine: Problems and Opportunities.'' In
\textit{Judgement Under Uncertainty: Heuristics and Biases}, edited by
Daniel Kahneman, Paul Slovic, and Amos Tversky. Cambridge University
Press, 1982.}

{
 Efran, M. G., and E. W. J. Patterson. ``The
Politics of Appearance.'' Unpublished PhD thesis,
1976.}

{
 Egan, Greg. \textit{Quarantine}. London: Legend Press, 1992.}

{
 Ehrlinger, Joyce, Thomas Gilovich, and Lee Ross.
``Peering Into the Bias Blind Spot:
People's Assessments of Bias in Themselves and
Others.'' \textit{Personality and Social Psychology
Bulletin} 31, no. 5 (2005): 680--692.}

{
 Eidelman, Scott, and Christian S. Crandall.
``Bias in Favor of the Status Quo.''
\textit{Social and Personality Psychology Compass} 6, no. 3 (2012):
270--281.}

{
 Epley, Nicholas, and Thomas Gilovich. ``Putting
Adjustment Back in the Anchoring and Adjustment Heuristic: Differential
Processing of Self-Generated and Experimentor-Provided
Anchors.'' \textit{Psychological Science} 12 (5
2001): 391--396. doi:10.1111/1467-9280.00372.}

{
 Epstein, Lewis Carroll. \textit{Thinking Physics: Understandable
Practical Reality, 3rd Edition}. Insight Press, 2009.}

{
 Feldman, Richard. ``Naturalized
Epistemology.'' In \textit{The Stanford Encyclopedia
of Philosophy}, Summer 2012, edited by Edward N. Zalta.}

{
 Festinger, Leon, Henry W. Riecken, and Stanley Schachter.
\textit{When Prophecy Fails: A Social and Psychological Study of a
Modern Group That Predicted the Destruction of the World}.
Harper-Torchbooks, 1956.}

{
 Fetherstonhaugh, David, Paul Slovic, Stephen M. Johnson, and James
Friedrich. ``Insensitivity to the Value of Human Life:
A Study of Psychophysical Numbing.'' \textit{Journal
of Risk and Uncertainty} 14, no. 3 (1997): 283--300.
doi:10.1023/A:1007744326393.}

{
 Feynman, Richard P. ``Judging Books by Their
Covers.'' In \textit{Surely You're
Joking, Mr. Feynman!} New York: W. W. Norton \& Company, 1985.}

{
 Feynman, Richard P., Robert B. Leighton, and Matthew L. Sands.
\textit{The Feynman Lectures on Physics}. 3 vols. Reading, MA:
Addison-Wesley, 1963.}

{
 Finucane, Melissa L., Ali Alhakami, Paul Slovic, and Stephen M.
Johnson. ``The Affect Heuristic in Judgments of Risks
and Benefits.'' \textit{Journal of Behavioral
Decision Making} 13, no. 1 (2000): 1--17.}

{
 Fong, Geoffrey T., David H. Krantz, and Richard E. Nisbett.
``The Effects of Statistical Training on Thinking
about Everyday Problems.'' \textit{Cognitive
Psychology} 18, no. 3 (1986): 253--292.
doi:10.1016/0010-0285(86)90001-0.}

{
 Frank, Adam. \textit{The Constant Fire: Beyond the Science vs.
Religion Debate}. University of California Press, 2009.}

{
 Freire, Paulo. \textit{The Politics of Education: Culture, Power,
and Liberation}. Greenwood Publishing Group, 1985.}

{
 Galbraith, John Kenneth. \textit{Economics, Peace and Laughter}.
Plume, 1981.}

{
 Ganzach, Yoav. ``Judging Risk and Return of
Financial Assets.'' \textit{Organizational Behavior
and Human Decision Processes} 83, no. 2 (2000): 353--370.
doi:10.1006/obhd.2000.2914.}

{
 Gendlin, Eugene T. \textit{Focusing}. Bantam Books, 1982.}

{
 Gibbon, Edward. \textit{The History of the Decline and Fall of the
Roman Empire}. J. \& J. Harper, 1829.}

{
 Gigerenzer, Gerd, and Ulrich Hoffrage. ``How to
Improve Bayesian Reasoning without Instruction: Frequency
Formats.'' \textit{Psychological Review} 102 (1995):
684--704.}

{
 Gilbert, Daniel T., Douglas S. Krull, and Patrick S. Malone.
``Unbelieving the Unbelievable: Some Problems in the
Rejection of False Information.'' \textit{Journal of
Personality and Social Psychology} 59 (4 1990): 601--613.
doi:10.1037/0022-3514.59.4.601.}

{
 Gilbert, Daniel T., and Patrick S. Malone. ``The
Correspondence Bias.'' \textit{Psychological
Bulletin} 117, no. 1 (1995): 21--38.
\url{http://www.wjh.harvard.edu/~dtg/Gilbert\%20\&\%20Malone\%20(CORRESPONDENCE\%20BIAS).pdf}.}

{
 Gilbert, Daniel T., Romin W. Tafarodi, and Patrick S. Malone.
``You Can't Not Believe Everything You
Read.'' \textit{Journal of Personality and Social
Psychology} 65 (2 1993): 221--233. doi:10.1037/0022-3514.65.2.221.}

{
 Gilbert, William S., and Arthur Sullivan. \textit{The Mikado}.
Opera, 1885.}

{
 Gilovich, Thomas, Dale Griffin, and Daniel Kahneman, eds.
\textit{Heuristics and Biases: The Psychology of Intuitive Judgment}.
New York: Cambridge University Press, 2002. doi:10.2277/0521796792.}

{
 Goertzel, Ben, and Cassio Pennachin, eds. \textit{Artificial
General Intelligence}. Cognitive Technologies. Berlin: Springer, 2007.
doi:10.1007/978-3-540-68677-4.}

{
 Good, Irving John. ``Speculations Concerning the
First Ultraintelligent Machine.'' In \textit{Advances
in Computers}, edited by Franz L. Alt and Morris Rubinoff, 6:31--88.
New York: Academic Press, 1965. doi:10.1016/S0065-2458(08)60418-0.}

{
 Graur, Dan, and Wen-Hsiung Li. \textit{Fundamentals of Molecular
Evolution}. 2nd ed. Sunderland, MA: Sinauer Associates, 2000.}

{
 Griffin, Dale, and Amos Tversky. ``The Weighing
of Evidence and the Determinants of Confidence.''
\textit{Cognitive Psychology} 24, no. 3 (1992): 411--435.
doi:10.1016/0010-0285(92)90013-R.}

{
 Haidt, Jonathan. ``The Emotional Dog and Its
Rational Tail: A Social Intuitionist Approach to Moral
Judgment.'' \textit{Psychological Review} 108, no. 4
(2001): 814--834. doi:10.1037/0033-295X.108.4.814.}

{
 Halberstadt, Jamin Brett, and Gary M. Levine.
``Effects of Reasons Analysis on the Accuracy of
Predicting Basketball Games.'' \textit{Journal of
Applied Social Psychology} 29, no. 3 (1999): 517--530.}

{
 Haldane, John B. S. ``A Mathematical Theory of
Natural and Artificial Selection.''
\textit{Mathematical Proceedings of the Cambridge Philosophical
Society} 23 (5 1927): 607--615. doi:10.1017/S0305004100011750.}

{
 Hamermesh, Daniel S., and Jeff E. Biddle.
``Beauty and the Labor Market.''
\textit{The American Economic Review} 84 (5 1994): 1174--1194.}

{
 Hansen, Katherine, Margaret Gerbasi, Alexander Todorov, Elliott
Kruse, and Emily Pronin. ``People Claim Objectivity
After Knowingly Using Biased Strategies.''
\textit{Personality and Social Psychology Bulletin} 40, no. 6 (2014):
691--699.}

{
 Hanson, Robin. ``You Are Never Entitled to Your
Opinion.'' \textit{Overcoming Bias (blog)} (2006).
\url{http://www.overcomingbias.com/2006/12/you\_are\_never\_e.html}.}

{
 Harris, Sam. \textit{The End of Faith: Religion, Terror, and the
Future of Reason}. WW Norton \& Company, 2005.}

{
 Hastorf, Albert, and Hadley Cantril. ``They Saw a
Game: A Case Study.'' \textit{Journal of Abnormal and
Social Psychology} 49 (1954): 129--134.
\url{http://www2.psych.ubc.ca/~schaller/Psyc590Readings/Hastorf1954.pdf}.}

{
 Heuer, Richards J. \textit{Psychology of Intelligence Analysis}.
Center for the Study of Intelligence, Central Intelligence Agency,
1999.}

{
 Hibbard, Bill. ``Super-Intelligent
Machines.'' \textit{ACM SIGGRAPH Computer Graphics}
35, no. 1 (2001): 13--15.
\url{http://www.siggraph.org/publications/newsletter/issues/v35/v35n1.pdf}.}

{
 Hintze, Daniel. ``Problem Class Dominance in
Predictive Dilemmas.'' Honors thesis (2014).}

{
 Hodgell, Patricia C. \textit{Seeker's Mask}.
Meisha Merlin Publishing, Inc., 2001.}

{
 Holt, Jim. ``Thinking Inside the
Boxes.'' \textit{Slate} (2002).
\url{http://www.slate.com/articles/arts/egghead/2002/02/thinkinginside\%5C\_the\%5C\_boxes.single.html}.}

{
 Holyoak, Keith J., and Robert G. Morrison. \textit{The Oxford
Handbook of Thinking and Reasoning}. Oxford University Press, 2013.}

{
 Hsee, Christopher K. ``Less Is Better: When
Low-Value Options Are Valued More Highly than High-Value
Options.'' \textit{Behavioral Decision Making} 11 (2
1998): 107--121.}

{
 {}---{}---{}---. ``The Evaluability Hypothesis:
An Explanation for Preference Reversals between Joint and Separate
Evaluations of Alternatives.'' \textit{Organizational
Behavior and Human Decision Processes} 67 (3 1996): 247--257.
doi:10.1006/obhd.1996.0077.}

{
 Hsee, Christopher K., and Howard C. Kunreuther.
``The Affection Effect in Insurance
Decisions.'' \textit{Journal of Risk and Uncertainty}
20 (2 2000): 141--159. doi:10.1023/A:1007876907268.}

{
 Hutter, Marcus. \textit{Universal Artificial Intelligence:
Sequential Decisions Based On Algorithmic Probability}. Berlin:
Springer, 2005. doi:10.1007/b138233.}

{
 Huxley, Thomas Henry. \textit{Evolution and Ethics and Other
Essays}. Macmillan, 1894.}

{
 Imagination Engines, Inc. ``The Imagination
Engine\textsuperscript{\textregistered}; or
Imagitron\textsuperscript{TM}.'' 2011.
\url{http://www.imagination-engines.com/ie.htm}.}

{
 Jaynes, Edwin T. \textit{Probability Theory: The Logic of
Science}. Edited by George Larry Bretthorst. New York: Cambridge
University Press, 2003. doi:10.2277/0521592712.}

{
 {}---{}---{}---. ``Probability Theory as
Logic.'' In \textit{Maximum Entropy and Bayesian
Methods}, edited by Paul F. Fougère. Springer Netherlands, 1990.}

{
 {}---{}---{}---. ``Probability Theory, with
Applications in Science and Engineering.''
Unpublished manuscript (1974).}

{
 Jones, Edward E., and Victor A. Harris. ``The
Attribution of Attitudes.'' \textit{Journal of
Experimental Social Psychology} 3 (1967): 1--24.
\url{http://www.radford.edu/~jaspelme/443/spring-2007/Articles/Jones\_n\_Harris\_1967.pdf}.}

{
 Joyce, James M. \textit{The Foundations of Causal Decision
Theory}. New York: Cambridge University Press, 1999.
doi:10.1017/CBO9780511498497.}

{
 Kahneman, Daniel. ``Comments by Professor Daniel
Kahneman.'' In \textit{Valuing Environmental Goods:
An Assessment of the Contingent Valuation Method}, edited by Ronald G.
Cummings, David S. Brookshire, and William D. Schulze, 1.B:226--235.
Experimental Methods for Assessing Environmental Benefits. Totowa, NJ:
Rowman \& Allanheld, 1986.
\url{http://yosemite.epa.gov/ee/epa/eerm.nsf/vwAN/EE-0280B-04.pdf}.}

{
 Kahneman, Daniel, Ilana Ritov, and David Schkade.
``Economic Preferences or Attitude Expressions?: An
Analysis of Dollar Responses to Public Issues.''
\textit{Journal of Risk and Uncertainty} 19, nos. 1--3 (1999):
203--235. doi:10.1023/A:1007835629236.}

{
 Kahneman, Daniel, David A. Schkade, and Cass R. Sunstein.
``Shared Outrage and Erratic Awards: The Psychology of
Punitive Damages.'' \textit{Journal of Risk and
Uncertainty} 16 (1 1998): 48--86. doi:10.1023/A:1007710408413.}

{
 Kahneman, Daniel, and Amos Tversky. ``Prospect
Theory: An Analysis of Decision Under Risk.''
\textit{Econometrica} 47 (1979): 263--292.}

{
 Keats, John. ``Lamia.''
\textit{The Poetical Works of John Keats} (London: Macmillan) (1884).}

{
 Keysar, Boaz. ``Language Users as Problem
Solvers: Just What Ambiguity Problem Do They Solve?''
In \textit{Social and Cognitive Approaches to Interpersonal
Communication}, edited by Susan R. Fussell and Roger J. Kreuz,
175--200. Mahwah, NJ: Lawrence Erlbaum Associates, 1998.}

{
 {}---{}---{}---. ``The Illusory Transparency of
Intention: Linguistic Perspective Taking in Text.''
\textit{Cognitive Psychology} 26 (2 1994): 165--208.
doi:10.1006/cogp.1994.1006.}

{
 Keysar, Boaz, and Dale J. Barr. ``Self-Anchoring
in Conversation: Why Language Users Do Not Do What They
`Should.'\,'' In
Gilovich, Griffin, and Kahneman, \textit{Heuristics and Biases},
150--166.}

{
 Keysar, Boaz, and Bridget Bly. ``Intuitions of
the Transparency of Idioms: Can One Keep a Secret by Spilling the
Beans?'' \textit{Journal of Memory and Language} 34
(1 1995): 89--109. doi:10.1006/jmla.1995.1005.}

{
 Keysar, Boaz, and Anne S. Henly.
``Speakers' Overestimation of Their
Effectiveness.'' \textit{Psychological Science} 13 (3
2002): 207--212. doi:10.1111/1467-9280.00439.}

{
 Kirk, Robert. \textit{Mind and Body}.
McGill-Queen's University Press, 2003.}

{
 Knishinsky, A. ``The Effects of Scarcity of
Material and Exclusivity of Information on Industrial Buyer Perceived
Risk in Provoking a Purchase Decision.'' Doctoral
dissertation, Arizona State University, 1982.}

{
 Kosslyn, Stephen M. \textit{Image and Brain: The Resolution of the
Imagery Debate}. Cambridge, MA: MIT Press, 1994.}

{
 Krips, Henry. ``Measurement in Quantum
Theory.'' In \textit{The Stanford Encyclopedia of
Philosophy}, Fall 2013, edited by Edward N. Zalta.}

{
 Kruschke, John K. \textit{Doing Bayesian Data Analysis, Second
Edition: A Tutorial with R, JAGS, and Stan}. Academic Press, 2014.}

{
 {}---{}---{}---. ``What to Believe: Bayesian
Methods for Data Analysis.'' \textit{Trends in
Cognitive Sciences} 14, no. 7 (2010): 293--300.}

{
 Kulka, Richard A., and Joan B. Kessler. ``Is
Justice Really Blind?: The Effect of Litigant Physical Attractiveness
on Judicial Judgment.'' \textit{Journal of Applied
Social Psychology} 8 (4 1978): 366--381.
doi:10.1111/j.1559-1816.1978.tb00790.x.}

{
 Kunreuther, Howard, Robin Hogarth, and Jacqueline Meszaros.
``Insurer Ambiguity and Market
Failure.'' \textit{Journal of Risk and Uncertainty} 7
(1 1993): 71--87. doi:10.1007/BF01065315.}

{
 Lakoff, George. \textit{Women, Fire, and Dangerous Things: What
Categories Reveal about the Mind}. Chicago: Chicago University Press,
1987.}

{
 Latané, Bibb, and John M. Darley. ``Bystander
`Apathy.'\,''
\textit{American Scientist} 57, no. 2 (1969): 244--268.
\url{http://www.jstor.org/stable/27828530}.}

{
 Lazarsfeld, Paul F. ``The American Solidier---An
Expository Review.'' \textit{Public Opinion
Quarterly} 13, no. 3 (1949): 377--404.}

{
 Le Guin, Ursula K. \textit{The Farthest Shore}. Saga Press, 2001.}

{
 Ledwig, Marion. ``Newcomb's
Problem.'' PhD diss., University of Constance, 2000.}

{
 Lichtenstein, Sarah, and Paul Slovic. ``Reversals
of Preference Between Bids and Choices in Gambing
Decisions.'' \textit{Journal of Experimental
Psychology} 89, no. 1 (1971): 46--55.}

{
 {}---{}---{}---, eds. \textit{The Construction of Preference}.
Cambridge University Press, 2006.}

{
 Lichtenstein, Sarah, Paul Slovic, Baruch Fischhoff, Mark Layman,
and Barbara Combs. ``Judged Frequency of Lethal
Events.'' \textit{Journal of Experimental Psychology:
Human Learning and Memory} 4, no. 6 (1978): 551--578.
doi:10.1037/0278-7393.4.6.551.}

{
 Liersch, Michael J., and Craig R. M. McKenzie.
``Duration Neglect by Numbers and Its Elimination by
Graphs.'' \textit{Organizational Behavior and Human
Decision Processes} 108, no. 2 (2009): 303--314.}

{
 Mack, Denise, and David Rainey. ``Female
Applicants' Grooming and Personnel
Selection.'' \textit{Journal of Social Behavior and
Personality} 5 (5 1990): 399--407.}

{
 MacKay, David J. C. \textit{Information Theory, Inference, and
Learning Algorithms}. New York: Cambridge University Press, 2003.}

{
 Matthews, Robert. ``Do We Need to Change the
Definition of Science?'' \textit{New Scientist} (May
2008).}

{
 Mazis, Michael B. ``Antipollution Measures and
Psychological Reactance Theory: A Field Experiment.''
\textit{Journal of Personality and Social Psychology} 31 (1975):
654--666.}

{
 Mazis, Michael B., Robert B. Settle, and Dennis C. Leslie.
``Elimination of Phosphate Detergents and
Psychological Reactance.'' \textit{Journal of
Marketing Research} 10 (1973): 390--395.}

{
 McDermott, Drew. ``Artificial Intelligence Meets
Natural Stupidity.'' \textit{SIGART Newsletter}, no.
57 (1976): 4--9. doi:10.1145/1045339.1045340.}

{
 McFadden, Daniel L., and Gregory K. Leonard.
``Issues in the Contingent Valuation of Environmental
Goods: Methodologies for Data Collection and
Analysis.'' In \textit{Contingent Valuation: A
Critical Assessment}, edited by Jerry A. Hausman, 165--215.
Contributions to Economic Analysis220. New York: North-Holland, 1993.
doi:10.1108/S0573-8555(1993)0000220007.}

{
 Mercier, Hugo, and Dan Sperber. ``Why Do Humans
Reason? Arguments for an Argumentative Theory.''
\textit{Behavioral and Brain Sciences} 34 (2011): 57--74.
\url{https://hal.archives-ouvertes.fr/file/index/docid/904097/filename/MercierSperberWhydohumansreason.pdf}.}

{
 Meyers, David G. \textit{Exploring Social Psychology}. New York:
McGraw-Hill, 1994.}

{
 Mitchell, Tom M. \textit{Machine Learning}. McGraw-Hill
Science/Engineering/Math, 1997.}

{
 Moore, John. \textit{Slay and Rescue}. Xlibris Corp, 2000.}

{
 Muehlhauser, Luke. ``The Power of
Agency.'' \textit{Less Wrong (blog)} (2011).
\url{http://lesswrong.com/lw/5i8/the\_power\_of\_agency/}.}

{
 Musashi, Miyamoto. \textit{Book of Five Rings}. New Line
Publishing, 2003.}

{
 Mussweiler, Thomas, and Fritz Strack. ``Comparing
Is Believing: A Selective Accessibility Model of Judgmental
Anchoring.'' \textit{European Review of Social
Psychology} 10 (1 1999): 135--167. doi:10.1080/14792779943000044.}

{
 Nagel, Thomas. ``What Is It Like to Be a
Bat?'' \textit{Philosophical Review} 83, no. 4
(1974): 435--450. \url{http://www.jstor.org/stable/2183914}.}

{
 Newby-Clark, Ian R., Michael Ross, Roger Buehler, Derek J.
Koehler, and Dale Griffin. ``People Focus on
Optimistic Scenarios and Disregard Pessimistic Scenarios While
Predicting Task Completion Times.'' \textit{Journal
of Experimental Psychology: Applied} 6, no. 3 (2000): 171--182.
doi:10.1037/1076-898X.6.3.171.}

{
 Nickerson, Raymond S. ``Confirmation Bias: A
Ubiquitous Phenomenon in Many Guises.''
\textit{Review of General Psychology} 2, no. 2 (1998): 175.}

{
 Nisbett, Richard E., and Timothy D. Wilson.
``Telling More than We Can Know: Verbal Reports on
Mental Processes.'' \textit{Psychological Review} 84
(1977): 231--259.
\url{http://people.virginia.edu/~tdw/nisbett\&wilson.pdf}.}

{
 Orwell, George. \textit{1984}. Signet Classic, 1950.}

{
 {}---{}---{}---. ``Politics and the English
Language.'' \textit{Horizon} (April 1946).}

{
 {}---{}---{}---. ``Why Socialists
Don't Believe in Fun.''
\textit{Tribune} (December 1943).}

{
 Pearce, David. \textit{The Hedonistic Imperative}.
\url{http://www.hedweb.com/}, 1995.}

{
 Pearl, Judea. \textit{Causality: Models, Reasoning, and
Inference}. 2nd ed. New York: Cambridge University Press, 2009.}

{
 {}---{}---{}---. \textit{Probabilistic Reasoning in Intelligent
Systems: Networks of Plausible Inference}. San Mateo, CA: Morgan
Kaufmann, 1988.}

{
 Peirce, Charles Sanders. \textit{Collected Papers}. Harvard
University Press, 1931.}

{
 Pinker, Steven. \textit{The Blank Slate: The Modern Denial of
Human Nature}. New York: Viking, 2002.}

{
 Piper, Henry Beam. ``Police
Operation.'' \textit{Astounding Science Fiction}
(July 1948).}

{
 Pirsig, Robert M. \textit{Zen and the Art of Motorcycle
Maintenance: An Inquiry Into Values}. 1st ed. New York: Morrow, 1974.}

{
 Planck, Max. \textit{Scientific Autobiography and Other Papers}.
New York: Philosophical Library, 1949.}

{
 Plato. \textit{Great Dialogues of Plato}. Edited by Eric H.
Warmington and Philip G. Rouse. Signet Classic, 1999.}

{
 Popper, Karl R. \textit{The Logic of Scientific Discovery}. New
York: Basic Books, 1959.}

{
 Poundstone, William. \textit{Priceless: The Myth of Fair Value
(and How to Take Advantage of It)}. Hill \& Wang, 2010.}

{
 Pratchett, Terry. \textit{Maskerade}. Discworld Series. ISIS,
1997.}

{
 {}---{}---{}---. \textit{Witches Abroad}. London: Corgi Books,
1992.}

{
 Procopius. \textit{History of the Wars}. Edited by Henry B.
Dewing. Harvard University Press, 1914.}

{
 Pronin, Emily. ``How We See Ourselves and How We
See Others.'' \textit{Science} 320 (2008):
1177--1180.
\url{http://psych.princeton.edu/psychology/research/pronin/pubs/2008\%20Self\%20and\%20Other.pdf}.}

{
 Pronin, Emily, Daniel Y. Lin, and Lee Ross. ``The
Bias Blind Spot: Perceptions of Bias in Self versus
Others.'' \textit{Personality and Social Psychology
Bulletin} 28, no. 3 (2002): 369--381.}

{
 Putnam, Hilary. ``The Meaning of
Meaning.'' In \textit{The Twin Earth Chronicles},
edited by Andrew Pessin and Sanford Goldberg, 3--52. M. E. Sharpe,
Inc., 1996.}

{
 Quattrone, George A., Cheryl P. Lawrence, Steven E. Finkel, and
David C. Andrus. ``Explorations in Anchoring: The
Effects of Prior Range, Anchor Extremity, and Suggestive
Hints.'' Unpublished manuscript, Stanford University,
1981.}

{
 Rhodes, Richard. \textit{The Making of the Atomic Bomb}. New York:
Simon \& Schuster, 1986.}

{
 Rips, Lance J. ``Inductive Judgments about
Natural Categories.'' \textit{Journal of Verbal
Learning and Verbal Behavior} 14 (1975): 665--681.}

{
 Roberts, Seth. ``What Makes Food Fattening?: A
Pavlovian Theory of Weight Control.'' Unpublished
manuscript, 2005.
\url{http://media.sethroberts.net/about/whatmakesfoodfattening.pdf}.}

{
 Robinson, Dave, and Judy Groves. \textit{Philosophy for
Beginners}. 1st ed. Cambridge: Icon Books, 1998.}

{
 Rorty, Richard. ``Out of the Matrix: How the Late
Philosopher Donald Davidson Showed That Reality Can't
Be an Illusion.'' \textit{The Boston Globe} (October
2003).}

{
 Rosch, Eleanor. ``Principles of
Categorization.'' In \textit{Cognition and
Categorization}, edited by Eleanor Rosch and Barbara B. Lloyd.
Hillsdale, NJ: Lawrence Erlbaum, 1978.}

{
 Russell, Gillian. ``Epistemic Viciousness in the
Martial Arts.'' In \textit{Martial Arts and
Philosophy: Beating and Nothingness}, edited by Graham Priest and Damon
A. Young. Open Court, 2010.}

{
 Russell, Stuart J., and Peter Norvig. \textit{Artificial
Intelligence: A Modern Approach}. 3rd ed. Upper Saddle River, NJ:
Prentice-Hall, 2010.}

{
 Ryle, Gilbert. \textit{The Concept of Mind}. University of Chicago
Press, 1949.}

{
 Sadock, Jerrold. ``Truth and
Approximations.'' \textit{Papers from the Third
Annual Meeting of the Berkeley Linguistics Society} (1977): 430--439.}

{
 Sagan, Carl. \textit{The Demon-Haunted World: Science as a Candle
in the Dark}. 1st ed. New York: Random House, 1995.}

{
 Saint-Exupery, Antoine de. \textit{Terre des Hommes}. Paris:
Gallimard, 1939.}

{
 Sandel, Michael. ``What's Wrong
With Enhancement.'' Background material for the
President's Council on Bioethics. (2002).}

{
 Schoemaker, Paul J. H. ``The Role of Statistical
Knowledge in Gambling Decisions: Moment vs. Risk Dimension
Approaches.'' \textit{Organizational Behavior and
Human Performance} 24, no. 1 (1979): 1--17.}

{
 Schul, Yaacov, and Ruth Mayo. ``Searching for
Certainty in an Uncertain World: The Difficulty of Giving Up the
Experiential for the Rational Mode of Thinking.''
\textit{Journal of Behavioral Decision Making} 16, no. 2 (2003):
93--106. doi:10.1002/bdm.434.}

{
 Schwitzgebel, Eric. \textit{Perplexities of Consciousness}. MIT
Press, 2011.}

{
 Serfas, Sebastian. \textit{Cognitive Biases in the Capital
Investment Context: Theoretical Considerations and Empirical
Experiments on Violations of Normative Rationality}. Springer, 2010.}

{
 Sherif, Muzafer, Oliver J. Harvey, B. Jack White, William R. Hood,
and Carolyn W. Sherif. ``Study of Positive and
Negative Intergroup Attitudes Between Experimentally Produced Groups:
Robbers Cave Study.'' Unpublished manuscript (1954).}

{
 Shermer, Michael. ``The Unlikeliest Cult in
History.'' \textit{Skeptic} 2, no. 2 (1993): 74--81.
\url{http://www.2think.org/02\_2\_she.shtml}.}

{
 Shimabukuro, Masayuki. \textit{Flashing Steel: Mastering
Eishin-Ryu Swordsmanship}. Frog Books, 1995.}

{
 Slovic, Paul. ``Numbed by
Numbers.'' \textit{Foreign Policy} (March 2007).
\url{http://foreignpolicy.com/2007/03/13/numbed-by-numbers/}.}

{
 Slovic, Paul, Melissa Finucane, Ellen Peters, and Donald G.
MacGregor. ``Rational Actors or Rational Fools:
Implications of the Affect Heuristic for Behavioral
Economics.'' \textit{Journal of Socio-Economics} 31,
no. 4 (2002): 329--342. doi:10.1016/S1053-5357(02)00174-9.}

{
 Smith, Edward Elmer. \textit{Second Stage Lensmen}. Old Earth
Books, 1998.}

{
 Smith, Edward Elmer, and Ric Binkley. \textit{Gray Lensman}. Old
Earth Books, 1998.}

{
 Smith, Edward Elmer, and A. J. Donnell. \textit{First Lensman}.
Old Earth Books, 1997.}

{
 Smullyan, Raymond M. \textit{What Is the Name of This Book?: The
Riddle of Dracula and Other Logical Puzzles}. Penguin Books, 1990.}

{
 Soares, Nate, and Benja Fallenstein. ``Toward
Idealized Decision Theory.'' Technical report.
Berkeley, CA: Machine Intelligence Research Institute (2014).
\url{http://intelligence.org/files/TowardIdealizedDecisionTheory.pdf}.}

{
 Spee, Friedrich. \textit{Cautio Criminalis; or, A Book on Witch
Trials}. Edited and translated by Marcus Hellyer. Studies in Early
Modern German History. 1631. Charlottesville: University of Virginia
Press, 2003.}

{
 Stanovich, Keith E., and Richard F. West.
``Individual Differences in Reasoning: Implications
for the Rationality Debate?'' \textit{Behavioral and
Brain Sciences} 23, no. 5 (2000): 645--665.
\url{http://journals.cambridge.org/abstract\_S0140525X00003435}.}

{
 Stewart, John E. ``Defendants'
Attractiveness as a Factor in the Outcome of Trials: An Observational
Study.'' \textit{Journal of Applied Social
Psychology} 10 (4 1980): 348--361.
doi:10.1111/j.1559-1816.1980.tb00715.x.}

{
 Stiegler, Marc. \textit{David's Sling}. Baen,
1988.}

{
 Stove, David Charles. \textit{The Plato Cult and Other
Philosophical Follies}. Cambridge University Press, 1991.}

{
 Strack, Fritz, and Thomas Mussweiler.
``Explaining the Enigmatic Anchoring Effect:
Mechanisms of Selective Accessibility.''
\textit{Journal of Personality and Social Psychology} 73, no. 3 (1997):
437--446.}

{
 Sunstein, Cass R. ``Probability Neglect:
Emotions, Worst Cases, and Law.'' \textit{Yale Law
Journal} (2002): 61--107.}

{
 Taber, Charles S., and Milton Lodge. ``Motivated
Skepticism in the Evaluation of Political Beliefs.''
\textit{American Journal of Political Science} 50, no. 3 (2006):
755--769. doi:10.1111/j.1540-5907.2006.00214.x.}

{
 Talbott, William. ``Bayesian
Epistemology.'' In \textit{The Stanford Encyclopedia
of Philosophy}, Fall 2013, edited by Edward N. Zalta.}

{
 TalkOrigins Foundation. ``Frequently Asked
Questions about Creationism and Evolution.''
\url{http://www.talkorigins.org/origins/faqs-qa.html}.}

{
 Tegmark, Max. \textit{Our Mathematical Universe: My Quest for the
Ultimate Nature of Reality}. Random House LLC, 2014.}

{
 {}---{}---{}---. ``Parallel
Universes.'' In \textit{Science and Ultimate Reality:
Quantum Theory, Cosmology, and Complexity}, edited by John D. Barrow,
Paul C. W. Davies, and Charles L. Harper Jr., 459--491. New York:
Cambridge University Press, 2004.}

{
 Thompson, Silvanus Phillips. \textit{The Life of Lord Kelvin}.
American Mathematical Society, 2005.}

{
 Thornton, Stephen. ``Karl
Popper.'' In \textit{The Stanford Encyclopedia of
Philosophy}, Winter 2002, edited by Edward N. Zalta. Stanford
University.
\url{http://plato.stanford.edu/archives/win2002/entries/popper/}.}

{
 Tooby, John, and Leda Cosmides. ``The
Psychological Foundations of Culture.'' In
\textit{The Adapted Mind: Evolutionary Psychology and the Generation of
Culture}, edited by Jerome H. Barkow, Leda Cosmides, and John Tooby,
19--136. New York: Oxford University Press, 1992.}

{
 Tversky, Amos, and Ward Edwards. ``Information
versus Reward in Binary Choices.'' \textit{Journal of
Experimental Psychology} 71, no. 5 (1966): 680--683.
doi:10.1037/h0023123.}

{
 Tversky, Amos, and Itamar Gati. ``Studies of
Similarity.'' In \textit{Cognition and
Categorization}, edited by Eleanor Rosch and Barbara Lloyd, 79--98.
Hillsdale, NJ: Lawrence Erlbaum Associates, Inc., 1978.}

{
 Tversky, Amos, and Daniel Kahneman. ``Extensional
Versus Intuitive Reasoning: The Conjunction Fallacy in Probability
Judgment.'' \textit{Psychological Review} 90, no. 4
(1983): 293--315. doi:10.1037/0033-295X.90.4.293.}

{
 {}---{}---{}---. ``Judgment Under Uncertainty:
Heuristics and Biases.'' \textit{Science} 185, no.
4157 (1974): 1124--1131. doi:10.1126/science.185.4157.1124.}

{
 {}---{}---{}---. ``Judgments of and by
Representativeness.'' In \textit{Judgment Under
Uncertainty: Heuristics and Biases}, edited by Daniel Kahneman, Paul
Slovic, and Amos Tversky, 84--98. New York: Cambridge University Press,
1982.}

{
 Tzu, Sun. \textit{The Art of War}. Cloud Hands, Inc., 2004.}

{
 Uhlmann, Eric Luis, and Geoffrey L. Cohen.
``\,`I think it, therefore
it's true': Effects of Self-perceived
Objectivity on Hiring Discrimination.''
\textit{Organizational Behavior and Human Decision Processes} 104, no.
2 (2007): 207--223.}

{
 Vaidman, Lev. ``Many-Worlds Interpretation of
Quantum Mechanics.'' In \textit{The Stanford
Encyclopedia of Philosophy}, Fall 2008, edited by Edward N. Zalta.}

{
 Vallone, Robert P., Lee Ross, and Mark R. Lepper.
``The Hostile Media Phenomenon: Biased Perception and
Perceptions of Media Bias in Coverage of the Beirut
Massacre.'' \textit{Journal of Personality and Social
Psychology} 49 (1985): 577--585.
\url{http://ssc.wisc.edu/~jpiliavi/965/hwang.pdf}.}

{
 Verhagen, Joachim.
\url{http://web.archive.org/web/20060424082937/http://www.nvon.nl/scheik/best/diversen/scijokes/scijokes.txt}.
Archived version, October 27, 2001.}

{
 Wade, Michael J. ``Group selections among
laboratory populations of Tribolium.''
\textit{Proceedings of the National Academy of Sciences of the United
States of America} 73, no. 12 (1976): 4604--4607.
doi:10.1073/pnas.73.12.4604.}

{
 Wansink, Brian, Robert J. Kent, and Stephen J. Hoch.
``An Anchoring and Adjustment Model of Purchase
Quantity Decisions.'' \textit{Journal of Marketing
Research} 35, no. 1 (1998): 71--81.
\url{http://www.jstor.org/stable/3151931}.}

{
 Warren, Adam. \textit{Empowered}. Dark Horse Books, 2007.}

{
 Wason, Peter Cathcart. ``On the Failure to
Eliminate Hypotheses in a Conceptual Task.''
\textit{Quarterly Journal of Experimental Psychology} 12, no. 3 (1960):
129--140. doi:10.1080/17470216008416717.}

{
 Weiten, Wayne. \textit{Psychology: Themes and Variations, Briefer
Version, Eighth Edition}. Cengage Learning, 2010.}

{
 West, Richard F., Russell J. Meserve, and Keith E. Stanovich.
``Cognitive Sophistication Does Not Attenuate the Bias
Blind Spot.'' \textit{Journal of Personality and
Social Psychology} 103, no. 3 (2012): 506.}

{
 Williams, George C. \textit{Adaptation and Natural Selection: A
Critique of Some Current Evolutionary Thought}. Princeton Science
Library. Princeton, NJ: Princeton University Press, 1966.}

{
 Wilson, David Sloan. ``A Theory of Group
Selection.'' \textit{Proceedings of the National
Academy of Sciences of the United States of America} 72, no. 1 (1975):
143--146.}

{
 Wilson, Timothy D., David B. Centerbar, and Nancy Brekke.
``Mental Contamination and the Debiasing
Problem.'' In \textit{Heuristics and Biases: The
Psychology of Intuitive Judgment}, edited by Thomas Gilovich, Dale
Griffin, and Daniel Kahneman. Cambridge University Press, 2002.}

{
 Wilson, Timothy D., Douglas J. Lisle, Jonathan W. Schooler, Sara
D. Hodges, Kristen J. Klaaren, and Suzanne J. LaFleur.
``Introspecting About Reasons Can Reduce Post-choice
Satisfaction.'' \textit{Personality and Social
Psychology Bulletin} 19 (1993): 331--331.}

{
 Wittgenstein, Ludwig. \textit{Philosophical Investigations}.
Translated by Gertrude E. M. Anscombe. Oxford: Blackwell, 1953.}

{
 Wright, Robert. \textit{The Moral Animal: Why We Are the Way We
Are: The New Science of Evolutionary Psychology}. Pantheon Books,
1994.}

{
 Yamagishi, Kimihiko. ``When a 12.86\% Mortality
Is More Dangerous than 24.14\%: Implications for Risk
Communication.'' \textit{Applied Cognitive
Psychology} 11 (6 1997): 461--554.}

{
 Yates, J. Frank, Ju-Whei Lee, Winston R. Sieck, Incheol Choi, and
Paul C. Price. ``Probability Judgment Across
Cultures.'' In Gilovich, Griffin, and Kahneman,
\textit{Heuristics and Biases}, 271--291.}

{
 Yudkowsky, Eliezer. ``Artificial Intelligence as
a Positive and Negative Factor in Global Risk.'' In
Bostrom and \'Cirkovi\'c, \textit{Global Catastrophic Risks},
308--345.}

{
 {}---{}---{}---. ``Cognitive Biases Potentially
Affecting Judgment of Global Risks.'' In Bostrom and
\'Cirkovi\'c, \textit{Global Catastrophic Risks}, 91--119.}

{
 {}---{}---{}---. \textit{Timeless Decision Theory}. Unpublished
manuscript. Machine Intelligence Research Institute, Berkeley, CA,
2010. \url{http://intelligence.org/files/TDT.pdf}.}

{
 Yuzan, Daidoji, William Scott Wilson, Jack Vaughn, and Gary Miller
Haskins. \textit{Budoshoshinshu: The Warrior's Primer
of Daidoji Yuzan}. Black Belt Communications Inc., 1984.}

{
 Zapato, Lyle. ``Lord Kelvin
Quotations.'' 2008.
\url{http://zapatopi.net/kelvin/quotes/}.}

{
 Zelazny, Roger. \textit{Prince of Chaos}. Thorndike Press, 2001.}

{
 Zhuangzi and Thomas Merton. \textit{The Way of Chuang Tzu}. New
Directions Publishing, 1965.}

{
 Zhuangzi and Burton Watson. \textit{The Complete Works of
Zhuangzi}. Columbia University Press, 1968.}


\bigskip


\bigskip

\clearpage
\bigskip
